### 2025-02-11 paper/body_ecml.tex
methodology done

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically bellow and analytically for
the linear case in \textbf{?@sec-app-constraints} of the supplementary
appendix.

@@

>>>>

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically\footnote{For the proof, see the supplementary
  appendix.}:

\begin{theorem}[Protecting Immutable
Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}

Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
class densities with common diagonal covariance matrix
\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
immutable feature from the contrastive divergence penalty
Equation~\ref{eq-div} will result in lower classifier sensitivity to
that feature, provided that at least one other feature is mutable.

\end{theorem}

It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxy features that
remain unprotected. We discuss this limitation in
Section~\ref{sec-discussion}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Limited to classification models.
\item
  Proxy attributes of immutable features.
\end{enumerate}

### 2025-02-11 paper/preprint/paper.tex
methodology done

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically below. Under certain
assumptions this can also be shown to hold analytically
Section~\ref{sec-app-constraints}.

@@

\begin{itemize}
\tightlist
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\end{itemize}

@@

\subsection{Details on Encoding Mutability
Constraints}\label{sec-app-constraints}

In Section~\ref{sec-constraints} we explain that to avoid penalizing
implausibility that arises due to mutability constraints, we impose a
point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
argue in Section~\ref{sec-constraints} that this approach induces models
to be less sensitive to feature \(d\) vis-à-vis some other mutable
feature \(m\) and demonstrate this empirically in
Section~\ref{sec-experiments}. Below we derive this result analytically
for linear classifier.

\subsubsection{Linear Classifier}\label{linear-classifier}

Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
\(\mathbf{x} \in \mathbb{R}^D\). Let \(d_{\text{mtbl}}\) and
\(d_{\text{immtbl}}\) denote some mutable and immutable feature,
respectively. Suppose that
\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
class \(k\).

@@

\begin{equation}\phantomsection\label{eq-loglike}{
\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
}\end{equation}

By Equation~\ref{eq-loglike} we have the
\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
and \(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\),
where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
\(d\) for class \(k\).

@@

Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
has the net affect of decreasing the classifier's sensitivity to the
immutable feature relative to the mutable feature (i.e.~no change in
sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
for \(d_{\text{mtbl}}\)).

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, opacityback=0, rightrule=.15mm, leftrule=.75mm, colback=white, left=2mm, colframe=quarto-callout-warning-color-frame, opacitybacktitle=0.6, breakable, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomtitle=1mm]

@Cynthia, @Arie, let me know if you want to phrase the above in terms of
a theorem and proof. This is something I've so far shied away from, but
I can see how it might be appropriate here.

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically\footnote{For the proof, see the supplementary
  appendix.}:

\begin{theorem}[Protecting Immutable
Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}

Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
class densities with common diagonal covariance matrix
\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
immutable feature from the contrastive divergence penalty
Equation~\ref{eq-div} will result in lower classifier sensitivity to
that feature, provided that at least one other feature is mutable.

\end{theorem}

It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxy features that
remain unprotected. We discuss this limitation in
Section~\ref{sec-discussion}.

@@

\begin{itemize}
\tightlist
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(y^-\): The non-target class and also the index of non-the target
  class.
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\end{itemize}

@@

\subsection{Protecting Mutability Constraints with Linear
Classifiers}\label{sec-app-constraints}

In Section~\ref{sec-constraints} we explain that to avoid penalizing
implausibility that arises due to mutability constraints, we impose a
point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
argue in Section~\ref{sec-constraints} that this approach induces models
to be less sensitive to immutable features and demonstrate this
empirically in Section~\ref{sec-experiments}. Below we derive the
analytical results in Theorem~\ref{thm-mtblty}.

\begin{proof}
Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
and immutable feature, respectively. Suppose that
\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
class \(k\). In words, we assume that the immutable feature tends to
take lower values for samples in the non-target class \(y^-\) than in
the target class \(y^+\). We assume the opposite to hold for the mutable
feature.

@@

\begin{equation}\phantomsection\label{eq-loglike}{
\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
}\end{equation}

By independence of \(x_1,...,x_D\), the full log-likelihood ratio
decomposes into:

\begin{equation}\phantomsection\label{eq-loglike-decomp}{
\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
}\end{equation}

By Equation~\ref{eq-loglike-decomp} we have the
\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
and \(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\),
where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
\(d\) for class \(k\).

@@

Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
has the net affect of decreasing the classifier's sensitivity to the
immutable feature relative to the mutable feature (i.e.~no change in
sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
for \(d_{\text{mtbl}}\)).
\end{proof}

@@

\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, titlerule=0mm, breakable, toprule=.15mm, left=2mm, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, opacityback=0, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, colback=white]

@Cynthia, @Arie, I have tentatively phrased the above in terms of a
theorem and proof. This is something I've so far shied away from because
I feel a bit out of my depth when it comes to mathematical proofs. The
above makes intuitive sense to me, but I don't know for sure if it's
correct.

### 2025-02-11 paper/sections/method.qmd
methodology done

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically bellow and analytically for the linear case in @sec-app-constraints of the supplementary appendix.

>>>>

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically^[For the proof, see the supplementary appendix.]:

::: {#thm-mtblty}

## Protecting Immutable Features

Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty [@eq-div] will result in lower classifier sensitivity to that feature, provided that at least one other feature is mutable.

It is worth highlighting that @thm-mtblty assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected. We discuss this limitation in @sec-discussion.

### 2025-02-11 paper/sections/other/constraints.qmd
methodology done

## Details on Encoding Mutability Constraints {#sec-app-constraints}

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to feature $d$ vis-à-vis some other mutable feature $m$ and demonstrate this empirically in @sec-experiments. Below we derive this result analytically for linear classifier.

### Linear Classifier

Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. Let $d_{\text{mtbl}}$ and $d_{\text{immtbl}}$ denote some mutable and immutable feature, respectively. Suppose that $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ and $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$, where $\mu_{k,d}$ denotes the conditional sample mean of feature $d$ in class $k$.

@@

$$
\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
$$ {#eq-loglike}

By @eq-loglike we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.

@@

\@Cynthia, \@Arie, let me know if you want to phrase the above in terms of a theorem and proof. This is something I've so far shied away from, but I can see how it might be appropriate here.

>>>>

## Protecting Mutability Constraints with Linear Classifiers {#sec-app-constraints}

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments. Below we derive the analytical results in @thm-mtblty.

::: {.proof}

Let $d_{\text{mtbl}}$ and $d_{\text{immtbl}}$ denote some mutable and immutable feature, respectively. Suppose that $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ and $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$, where $\mu_{k,d}$ denotes the conditional sample mean of feature $d$ in class $k$. In words, we assume that the immutable feature tends to take lower values for samples in the non-target class $y^-$ than in the target class $y^+$. We assume the opposite to hold for the mutable feature.

@@

$$
\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
$$ {#eq-loglike}

By independence of $x_1,...,x_D$, the full log-likelihood ratio decomposes into:

$$
\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
$$ {#eq-loglike-decomp}

By @eq-loglike-decomp we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.

@@

\@Cynthia, \@Arie, I have tentatively phrased the above in terms of a theorem and proof. This is something I've so far shied away from because I feel a bit out of my depth when it comes to mathematical proofs. The above makes intuitive sense to me, but I don't know for sure if it's correct.
